---
title: "RCaN - a primer with 2 species"
author: "Benjamin Planque"
date: "2021"
output: 
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A Primer for running Chance and Necessity (CaN) models in R

Benjamin Planque - 2021

*Institute of Marine Research. P.O, Box 6606 Langnes, 9296 Tromsoe, Norway*

*Tel: +47 48 89 30 43, email: benjamin.planque@hi.no, www.hi.no/en*

RCaN is developped and maintained by Hilaire Drouineau and available from:
https://github.com/inrae/RCaN

## STEP 1 - Install and load librairies

```{r installation & loading of librairies, echo=TRUE, results='hide'}
#install.packages("symengine")
#devtools::install_github("https://github.com/inrae/RCaN.git", subdir="RCaN")
#install.packages("ggplot2")
#install.packages("ggraph")
#install.packages("coda")
#install.packages("DT")
#install.packages("tidyverse")

require(symengine) # Symbolic math
require(RCaN) # the RCaN library
require(ggplot2) # plots
require(ggraph) # network plots
require(coda) # mcmc plotting
require(DT) # to visualise data tables
require(tidyverse) # get access to tidy syntax
```

## STEP 2 - load a RCaN file and build the corresponding model

This is a simple model with two populations. The first feeds on an external food source. The second population predates on the first one.

There are three explicit contraints:<br>
1. the feeding rate should not exceed 100 <br>
2. the biomass of population1 should not exceed observations*1.2 <br>
3. the biomass of population1 should not be below observations/1.2 <br>
To begin with, only the first constraint is active

There several implicit constraints <br>
1. the biomass of the populations should not go below the refuge biomass parameter <br>
2. the feeding rates can not exceed satiation (sigma*biomass) <br>
3. the relative change in population biomass from one year to the next is bounded by inertia (it cannot exceed exp(±mu))


```{r build model1, echo=TRUE}
rm(list=ls())
CaNfile <- "./TwoSpeciesRCaN.xlsx"
CaNmod <- buildCaN(CaNfile)

```

## STEP 3 - explore model elements

```{r exploring elements, echo=TRUE}
datatable(CaNmod$components_param[,1:8])
datatable(CaNmod$fluxes_def)
datatable(CaNmod$constraints)
datatable(CaNmod$series)
ggNetwork(CaNmod)
```

## STEP 4 - check that there can be solutions to this model
"polytope OK" indicates that this model has multiple solutions. This is the desired situation :-)

Other possibilities are  <br>
"empty polytope"  <br>
"polytope not bounded" <br> 
"unique solution"  <br>
"numerical error"  <br>
"potential problem" 

```{r check polytope status, echo=TRUE}
checkPolytopeStatus(CaNmod)
```
## STEP 4b - Check the bounds (maximum and minimum) for all flows in the model (and biomasses at time t=1)
```{r check polytope bounds, echo=TRUE,results='hide'}
gab <- getAllBoundsParam(CaNmod)
```
```{r show polytope bounds, echo=TRUE}
datatable(gab)
```

## STEP 5 - sample solutions to the model
We use three sampling sequences (chains) and draw 1000 samples in each of them. The method used is the Gibbs sampler and there is no thining, i.e. all samples are kept.
```{r sample the model, echo=TRUE, results='hide'}
CaNsample <- sampleCaN(CaNmod,1000, nchain = 3, ncore = 3, method = "gibbs")
```
## STEP 6 - explore the results
Plot the time series for the population <br>
Plot the time series for the feeding <br>
Check the sampling performance: are the distributions convex? are the chains autocorrelated?
```{r explore results, echo=TRUE}
Comp2plot <- CaNmod$species
ggSeries(CaNsample,param = Comp2plot, plot_series = TRUE, ylab = 'Biomasses (1000t)')
Flux2plot <- CaNmod$fluxes_def$Flux
ggSeries(CaNsample,param = Flux2plot, plot_series = TRUE, ylab = 'fluxes (1000t/year)')
CanNames <- attributes(CaNsample$mcmc[[1]])$dimnames[2]

for(i in 1:4){
  par(mfrow=c(2,2))
  hist(CaNsample$mcmc[,i][[1]],breaks = 25,main = CanNames[[1]][i])
  traceplot(CaNsample$mcmc[,i])
  plot(autocorr(CaNsample$mcmc[,i:i][[1]],lags=1:25),type='l')
  abline(h=0,col='red')
}
```

Everything looks great aside from the autocorrelation in the sampling. It might be worth "thining" the sampling, i.e. discarding samples.

## STEP 7 - revise the sampling
Same sampling as above, but this time we apply a thining of 50, i.e. we only retain one every 50 samples.
```{r sample the model 2, echo=TRUE, results='hide'}
CaNsampleThin <- sampleCaN(CaNmod,1000, nchain = 3, ncore = 3, method = "gibbs", thin = 50)
for(i in 1:4){
  par(mfrow=c(2,2))
  hist(CaNsampleThin$mcmc[,i][[1]],breaks = 25,main = CanNames[[1]][i])
  traceplot(CaNsampleThin$mcmc[,i])
  plot(autocorr(CaNsampleThin$mcmc[,i:i][[1]],lags=1:25),type='l')
  abline(h=0,col='red')
}
```

## STEP 8 - add constraints based on observations (C02 and C03)
We now include information for observational time series on the biomass of species 1. 
This is done by activating constraints C02 and C03 which state that the biomass trajectories sampled by RCaN should correspond to the observed biomasses /*1.2. <br>
How does constraining the model on population 1 alters population 2 and the trophic flows?


```{r toggle C03 and C04, echo=TRUE, results='hide'}
CaNmod2 <- toggleConstraint(CaNmod,c("C02","C03"))
datatable(CaNmod2$constraints)
checkPolytopeStatus(CaNmod2)
gab <- getAllBoundsParam(CaNmod2)
datatable(gab[1:10,])
CaNsample2 <- sampleCaN(CaNmod2,1000, nchain = 3, ncore = 3, method = "gibbs", thin = 50)
ggSeries(CaNsample2,param = Comp2plot, plot_series = TRUE, ylab = 'Biomasses (1000t)')
ggSeries(CaNsample2,param = Flux2plot, plot_series = TRUE, ylab = 'fluxes (1000t/year)')
```

```{r extra plot Flux1 vs Flux2, echo=FALSE}
  # CaNsampleLong <- as_tibble(as.matrix(CaNsample$mcmc)) %>%
  #   mutate(Sample_id = 1:dim(as.matrix(CaNsample$mcmc))[1]) %>%
  #   pivot_longer(col = -Sample_id,
  #                names_to = c("Var","Year"),
  #                names_pattern = "(.*)\\[(.*)\\]",
  #                values_to = 'value')
  # CaNsampleWide2005 <- pivot_wider(CaNsampleLong, names_from = "Var") %>%
  #   mutate(Year = as.numeric(Year)) %>%
  #   filter(Year == 2005)
  # ggplot(CaNsampleWide2005,aes(x=Food_Population1,y=Population1_Population2)) +
  #   geom_point()
```
